---
title: "GAPPS Phthalate WGCNA Figure"
author: "Samantha Lapehn & Mariana Parenti"
output: html_document
---

20251015 - Updates to improve visualizations: scale both placental efficiency measures. 

## Aims 

a) Generate the WGCNA dotplot, showing the WGCNA modules associated with the phthalate metabolites, DEHP, and the mixture, as well as BW~adj~ and BW:PW ratio. 
b) Generate the plot showing the enriched KEGG pathways for each WGCNA module identified in this analysis. 

```{r load packages}
library(tidyverse)
library(pheatmap)
library(here)
library(ggtext)
```

## Load  Results

```{r load results}
sig_PHT <- read_csv(here::here("Mariana", "Draft3", "MITM_WGCNA_Figure", "SigWGCNA_071025.csv"))[, -1]

BW_PE_modules <- read_csv(here::here("Mariana", "Env Int Revision", "output", "WGCNA_bw_ratio_fullResults.csv"))

mixtures_res <- read_csv(here::here("Mariana", "Draft3", "MITM_WGCNA_Figure", "GAPPS_Phthalate_WGCNA_QGComp_SigResults_071025.csv"))[, -1] |>
  ## select and rename columns
  dplyr::select(probeID, estimate = beta, std.error = se, statistic = tval, p.value = pval) |>
  ## format probeID as module
  dplyr::mutate(module = substring(probeID, 3)) |>
  dplyr::select(-probeID) |>
  dplyr::relocate(module) |>
  ## add mixture column
  dplyr::mutate(analysis = "Mixture") 

mixtures_res
```

## Prepare Data

```{r}
sig_PHT$phthalate <- toupper(sig_PHT$phthalate)
sig_PHT <- sig_PHT |>
  dplyr::rename(analysis = phthalate) |>
  dplyr::bind_rows(mixtures_res)

BW_PE_modules <- BW_PE_modules |>
  dplyr::filter(p.value<0.05)

All_modules <- dplyr::bind_rows(sig_PHT, BW_PE_modules)

All_modules <- All_modules |>
  mutate(analysis = ifelse(analysis == "BW~adj~", "BW(adj)", analysis))
```

## Make Summary Dot Plot

```{r DotPlot, fig.width=8, fig.height=8}
levels(as.factor(All_modules$module)) |> dput()

# Module <- c(black="black", blue="blue", brown="chocolate4", 
#             darkgreen="darkgreen", darkolivegreen="darkolivegreen", 
#             darkorange="darkorange3", darkturquoise="darkturquoise", 
#             green="green", greenyellow="olivedrab2", grey60="grey60", 
#             lightcyan="lightcyan3", lightgreen="seagreen2", magenta="magenta", 
#             orange="orange", purple="purple", red="red", royalblue= "royalblue",
#             salmon="salmon", sienna3="sienna3", skyblue="skyblue3", tan="tan",
#             turquoise="turquoise", violet="mediumpurple1", white="ivory3",
#             yellow="gold2",yellowgreen="yellowgreen")

Module <- c(black = "black", blue = "blue", brown = "chocolate4", cyan = "cyan3",
            darkgreen = "darkgreen", darkolivegreen = "darkolivegreen", 
            darkorange = "darkorange", darkturquoise = "darkturquoise",
            green = "green3", greenyellow = "olivedrab3", grey60 = "grey60", 
            lightcyan = "lightcyan3", lightgreen = "seagreen2",
            lightyellow = "lightgoldenrod2", magenta = "magenta", 
            orange = "orange", pink = "palevioletred1", red = "red", 
            royalblue = "royalblue", salmon = "salmon", sienna3 = "sienna3",
            skyblue = "skyblue", skyblue3 = "skyblue3", steelblue = "steelblue3",
            tan = "tan", turquoise = "turquoise", violet = "mediumpurple1", 
            white = "ivory3", yellow = "gold2", yellowgreen = "yellowgreen")

analysis_order <- c("BW:PW", "BW(adj)", "DEHP", "MBP", "MBZP", "MCHPP", "MCINP", 
                    "MCIOP", "MCMHP", "MCPP", "MECPP", "MEHHP", "MEHP", "MEOHP", 
                    "MEP", "MHPP", "MIBP", "MMP", "Mixture")
All_modules |> 
         dplyr::mutate(analysis = factor(analysis, levels = analysis_order)) |>
  ggplot() +
  aes(x=analysis, 
      y=module, 
      fill = estimate) + 
  ## labels
  labs(color = "Effect Estimate",
       size="Abs(Effect Estimate)",
       y = "Module", 
       x="Analysis") +
  ## data
  geom_point(
    aes(size = abs(estimate)),
    shape = 21, 
    color = "black", 
    stroke = 0.2) +
  theme_bw() + 
  ## scales
  scale_fill_gradient2(midpoint=0, 
                       low="#0092B3", 
                       mid="White", 
                       high="#F09062") +
  guides(size = guide_legend(override.aes = list(fill = "black"))) +
  ## themes
  theme_bw() + 
  theme(
    axis.text.x = element_markdown(angle=45, size=12, hjust=1),
    axis.text.y=element_text(size=12, color = Module, face = "bold"), 
    legend.text=element_text(size=12), 
    axis.title= element_text(size=14)) 

ggsave(filename = "MITM_WGCNA_dotplot_20250710.pdf", path = here::here("Mariana", "Env Int Revision", "output"), width = 8, height = 8, units = "in")
```

## Pathway Enrichment in the WGCNA modules

```{r fig.height=9, fig.width=5}
pathdata <- read_csv(here::here("Resubmission_July2024/output/wgcna/allPathwaysEnrichmentResults.csv")) |>
  dplyr::filter(FDR<0.2) |>
  dplyr::filter(module %in% All_modules$module) |>
  dplyr::mutate(magnitude = -log(P.DE),
                pathway_name = stringr::str_replace_all(Pathway, "_", " "),
                FDR_05_level = dplyr::if_else(FDR<0.05, 1, 0) |> factor()) |>
  dplyr::mutate(pathway_name = dplyr::if_else(
    pathway_name == "Hippo signaling pathway multiple species", 
    "Hippo signaling pathway - multiple species",
    pathway_name)) |>
  dplyr::mutate(pathway_name = dplyr::if_else(
    pathway_name == "Autophagy animal", 
    "Autophagy - animal",
    pathway_name)) 

levels(factor(pathdata$module)) |> dput()

library(jsonlite)

kegg_hierarchy <- jsonlite::read_json("https://rest.kegg.jp/get/br:br08901/json", 
                                      simplifyVector = FALSE) [[2]]

kegg_pathway_groups <- purrr::map_dfr(kegg_hierarchy,  ~ as_tibble(.x)) |>
  tidyr::unnest_wider(col = children, names_sep = "_") |>
  tidyr::unnest_wider(col = children_children, names_sep = "_") |>
  tidyr::pivot_longer(cols = !name:children_name,
                      names_to = "child_name",
                      values_to = "list") |>
  tidyr::unnest_wider(col = list, names_sep = "_") |>
  dplyr::select(supergroup = name, 
                subgroup = children_name, 
                id_pathway = list_name) |>
  dplyr::filter(is.na(id_pathway) == FALSE) |>
  dplyr::mutate(kegg_pathway_id = stringr::str_split_fixed(id_pathway, " ", 2)[, 1],
                pathway_name = stringr::str_split_fixed(id_pathway, " ", 2)[, 2]) |>
  dplyr::mutate(pathway_name= substring(pathway_name, 2))

pathway_order <- kegg_pathway_groups |>
  dplyr::filter(pathway_name %in% pathdata$pathway_name) |>
  dplyr::arrange(supergroup, subgroup, pathway_name) |>
  dplyr::pull(pathway_name)

pathway_plot_ordered <- pathdata |>
  dplyr::mutate(pathway_name = factor(pathway_name, levels = pathway_order)) |>
  ggplot() +
  aes(x = factor(module), 
      y = pathway_name) +
  geom_point(aes(size = magnitude, 
                 fill = module, 
                 shape = FDR_05_level),
             stroke = 0.2) +
  # geom_point(data = pathdata |> dplyr::filter(FDR <0.05),
  #            size = 1, stroke = 0.5,
  #            shape = 8, alpha = 0.5) +
  ylab("") +
  xlab("") +
  scale_fill_manual(values = c("blue", "brown", "cyan", "darkgreen", 
                               "darkorange", "darkturquoise", "green", 
                               "greenyellow", "lightcyan", "lightyellow", 
                               "magenta", "orange", "pink", "red", "royalblue",
                               "skyblue", "skyblue3", "steelblue", "tan", 
                               "turquoise", "yellow", "yellowgreen")) +
  scale_shape_manual(values = c(22, 21), name = "FDR", labels = c("<0.20", "<0.05")) +
  theme_bw(base_size = 10) +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.background = element_blank(), 
    plot.background = element_blank())

pathway_plot_ordered

ggsave(filename = here::here("Mariana", "Env Int Revision", "output", "wgcnaModulePathways_MITA_transparent.pdf"),
       width = 5.5, height = 11, units = "in")
```
